---
title: "Compare velociraptor results to results obtained directly with `scVelo`"
subtitle: "Slightly modified from an extra vignette contained in the velociraptor package"
author: "Charlotte Soneson"
date: "April 26, 2022"
output:
  rmarkdown::html_document:
    css: styles.css
    highlight: pygments
    toc: true
    toc_float: true
    toc_depth: 3
    fig_width: 5
    number_sections: false
    theme: united
bibliography: "rna-velocity.bib"
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE, class.source="rchunk"}
knitr::opts_chunk$set(echo = TRUE)
```

# Load packages

```{r, class.source="rchunk"}
suppressPackageStartupMessages({
    library(scRNAseq)
    library(scuttle)
    library(scran)
    library(velociraptor)
    library(scater)
    library(cowplot)
    library(GGally)
    library(reticulate)
    library(SingleCellExperiment)
    library(basilisk)
    library(BiocStyle)
})
```

# Define virtual environment to use for reticulate

We will use the `velociraptor` conda environment in order to run the _scVelo_ commands also directly in python.

```{r, class.source="rchunk"}
reticulate::use_condaenv(basilisk:::.obtainEnvironmentPath(velociraptor:::velo.env))
sys <- reticulate::import("sys")
sys$version
```

# Set temporary directory

Output objects (to move data from SCE to AnnData without using basilisk) will be saved to a temporary directory.

```{r, class.source="rchunk"}
tmpdir <- tempdir()
```

# Prepare data

We use an example data set from the `r Biocpkg("scRNAseq")` package, and subset to the first 500 cells. 
The object is also saved to a `.h5ad` file for use with _scVelo_ directly in python.

```{r, class.source="rchunk"}
## Load and subset data
sce <- scRNAseq::HermannSpermatogenesisData()[, 1:500]

## Add 'counts' assay
assay(sce, "counts") <- assay(sce, "spliced")

## Save to h5ad file for python scVelo analysis
zellkonverter::writeH5AD(sce, X_name = "counts", 
                         file = file.path(tmpdir, "anndata.h5ad"))
```

# Run analysis directly in python

We use the same conda environment as in `velociraptor` to run the _scVelo_ analysis directly in python, rather than via the R wrapper. 

## Steady-state model

```{python, class.source="pythonchunk"}
import scanpy as sc
import sys
import numpy as np
import anndata
import scvelo as scv
import matplotlib
import pandas as pd
from pathlib import Path
from os import path
adata=anndata.read(r.tmpdir + "/anndata.h5ad")
scv.pp.filter_and_normalize(adata, enforce=True, n_top_genes=2000)
scv.pp.moments(adata)
scv.tl.velocity(adata, mode="steady_state")
scv.tl.velocity_graph(adata)
scv.tl.velocity_pseudotime(adata)
scv.tl.velocity_confidence(adata)
adata.write(r.tmpdir + "/anndata_withvelo_steadystate.h5ad")
```

## Dynamical model

```{python, class.source="pythonchunk"}
import scanpy as sc
import sys
import numpy as np
import anndata
import scvelo as scv
import matplotlib
import pandas as pd
adata=anndata.read(r.tmpdir + "/anndata.h5ad")
scv.pp.filter_and_normalize(adata, enforce=True, n_top_genes=2000)
scv.pp.moments(adata)
scv.tl.recover_dynamics(adata)
scv.tl.velocity(adata, mode="dynamical")
scv.tl.velocity_graph(adata)
scv.tl.velocity_pseudotime(adata)
scv.tl.latent_time(adata)
scv.tl.velocity_confidence(adata)
adata.write(r.tmpdir + "/anndata_withvelo_dynamical.h5ad")
```

## Read back objects in R for later comparison

```{r, class.source="rchunk"}
## Steady-state model, python
sceps <- zellkonverter::readH5AD(file.path(tmpdir, "anndata_withvelo_steadystate.h5ad"))

## Dynamical model, python
scepd <- zellkonverter::readH5AD(file.path(tmpdir, "anndata_withvelo_dynamical.h5ad"))
```

# Process the data set

Next, we apply some basic preprocessing to the data set - normalization, selection of highly variable genes, and dimension reduction. 

```{r, class.source="rchunk"}
sce <- scuttle::logNormCounts(sce, assay.type=1)

## To use the HVGs determined by scVelo
# top.hvgs <- rownames(sceps)

## To get HVGs with scran
dec <- scran::modelGeneVar(sce)
top.hvgs <- scran::getTopHVGs(dec, n=2000)

set.seed(1)
sce <- scater::runPCA(sce, subset_row=top.hvgs)
sce <- scater::runTSNE(sce, dimred="PCA")
```

# Helper function

The function below runs scVelo via the R wrapper, first using the pre-normalized data and PCA from the SingleCellExperiment object, and then with `use_theirs=TRUE`. 
It also performs some comparisons between the two results. 

```{r, class.source="rchunk"}
run_scv <- function(sce, top.hvgs, mode) {
    ## precomputed PCs and normalized values
    velo.out <- scvelo(
        sce, assay.X="counts", sf.X=sizeFactors(sce), 
        subset.row=top.hvgs, use.dimred="PCA",
        mode=mode
    )
    stopifnot(colnames(sce) == colnames(velo.out))
    sce$velocity_pseudotime <- velo.out$velocity_pseudotime
    
    ## using scVelo normalization, gene selection and PCA calculation
    velo.out_theirs <- scvelo(
        sce, assay.X="counts", use.theirs=TRUE, mode=mode,
        scvelo.params=list(filter_and_normalize=list(enforce=TRUE, 
                                                     n_top_genes=2000L))
    )
    stopifnot(colnames(sce) == colnames(velo.out_theirs))
    sce$velocity_pseudotime_theirs <- velo.out_theirs$velocity_pseudotime

    ## Inferred pseudotime
    print(cowplot::plot_grid(
        cowplot::plot_grid(
            plotTSNE(sce, colour_by="velocity_pseudotime"),
            plotTSNE(sce, colour_by="velocity_pseudotime_theirs"),
            nrow = 1),
        plotTSNE(sce, colour_by="celltype"), ncol = 1
    ))
    print(GGally::ggpairs(as.data.frame(colData(sce))[, c("velocity_pseudotime", 
                                                          "velocity_pseudotime_theirs")]))

    shared_genes <- Reduce(intersect, list(rownames(velo.out), rownames(velo.out_theirs)))
    message("Number of shared genes:")
    print(length(shared_genes))
    message("Ours:")
    print(table(shared = rownames(velo.out) %in% shared_genes,
                velogene = rowData(velo.out)$velocity_genes))
    message("Theirs:")
    print(table(shared = rownames(velo.out_theirs) %in% shared_genes,
                velogene = rowData(velo.out_theirs)$velocity_genes))
    message("Among shared genes, how many are velocity genes:")
    print(table(ours = rowData(velo.out[shared_genes, ])$velocity_genes,
                theirs = rowData(velo.out_theirs[shared_genes, ])$velocity_genes))
    
    ## Correlations of velocities
    velo_corrs <- diag(cor(assay(velo.out[shared_genes, ], "velocity"), 
                           assay(velo.out_theirs[shared_genes, colnames(velo.out)],
                                 "velocity"),
                           use = "pairwise.complete"))
    hist(velo_corrs, 100)
    
    ## Correlations of normalized spliced/unspliced values
    ms_corrs <- diag(cor(assay(velo.out[shared_genes, ], "Ms"), 
                         assay(velo.out_theirs[shared_genes, colnames(velo.out)],
                               "Ms"),
                         use = "pairwise.complete"))
    hist(ms_corrs, 100)
    mu_corrs <- diag(cor(assay(velo.out[shared_genes, ], "Mu"), 
                         assay(velo.out_theirs[shared_genes, colnames(velo.out)],
                               "Mu"),
                         use = "pairwise.complete"))
    hist(mu_corrs, 100)
    
    velo.out_theirs
}
```

# Run velociraptor and compare to python output

## Dynamical model

```{r, fig.width = 8, class.source="rchunk"}
sced <- run_scv(sce, top.hvgs=top.hvgs, mode="dynamical")
```

### Compare velociraptor run with 'use_theirs=TRUE' to direct python/scVelo calls

```{r, class.source="rchunk"}
stopifnot(all(colnames(sced) == colnames(scepd)),
          all(rownames(sced) == rownames(scepd)))
## Correlations of velocities
velo_corrs <- diag(cor(assay(sced, "velocity"), 
                       assay(scepd, "velocity"),
                       use = "pairwise.complete"))
summary(velo_corrs)
plot(sced$velocity_pseudotime, scepd$velocity_pseudotime)
summary(abs(sced$velocity_pseudotime - scepd$velocity_pseudotime))
stopifnot(max(abs(sced$velocity_pseudotime - scepd$velocity_pseudotime)) < 1e-3)
```

## Steady-state model

```{r, fig.width = 8, class.source="rchunk"}
sces <- run_scv(sce, top.hvgs=top.hvgs, mode="steady_state")
```

### Compare velociraptor run with 'use_theirs=TRUE' to direct python/scVelo calls

```{r, class.source="rchunk"}
stopifnot(all(colnames(sces) == colnames(sceps)),
          all(rownames(sces) == rownames(sceps)))
## Correlations of velocities
velo_corrs <- diag(cor(assay(sces, "velocity"), 
                       assay(sceps, "velocity"),
                       use = "pairwise.complete"))
summary(velo_corrs)
plot(sces$velocity_pseudotime, sceps$velocity_pseudotime)
summary(abs(sces$velocity_pseudotime - sceps$velocity_pseudotime))
stopifnot(max(abs(sces$velocity_pseudotime - sceps$velocity_pseudotime)) < 1e-3)
```


# Session info

```{r, class.source="rchunk"}
sessionInfo()
```

