---
title: "Cluster-based Differential Abundance analysis"
author: "Pierre-Luc"
date: "4/11/2022"
output: html_document
---

```{r}
suppressPackageStartupMessages({
  library(SingleCellExperiment) # core sc infrastructures
  library(edgeR) # differential expression analysis
  library(ANCOMBC) # differential abundance analysis with composition bias
  library(phyloseq) # microbiome data structure, we'll need it for ANCOMBC
  library(ggplot2) # for plotting
  library(patchwork) # for combining plots
  library(ComplexHeatmap) # for plotting heatmaps
})
```

# Differential abundance analysis

## Using edgeR & co

The simplest approach is to use established count-based analysis methods (e.g. edgeR \& co).

We first load the data and extract the cluster counts per sample:

```{r}
#sce <- readRDS("/data/day1_differential_analysi/DA_example1.SCE.rds")
sce <- readRDS("../../DA_example1.SCE.rds")
# we build a SE of cluster abundances
cluster_abundances <- unclass(table(sce$cluster_id, sce$sample_id))
names(dimnames(cluster_abundances)) <- NULL
cla <- SummarizedExperiment(list(counts=as.matrix(cluster_abundances[,-1])))
cla$group <- factor(ifelse(grepl("LPS",colnames(cla)),"LPS","Control"))
# we calculate proportions
assays(cla)$proportion <- 100*t(t(assay(cla))/colSums(assay(cla)))
```

We plot the proportions across group:

```{r}
Heatmap(assays(cla)$proportion, column_split=cla$group, name="%")
# scaled proportion (to view smaller populations)
Heatmap(t(scale(t(assays(cla)$proportion))), column_split=cla$group, name="scaled %")

# quick fn to compare the proportions across groups
plotProps <- function(cla, assay="proportion"){
  # get the mean proportions per group
  m <- sapply(split(seq_len(ncol(cla)), cla$group), FUN=function(i){
    rowMeans(assays(cla)[[assay]][,i])
  })
  m <- cbind(cluster_id=row.names(m), as.data.frame(m))
  ggplot(m, aes(Control, LPS, colour=cluster_id)) + geom_abline() + 
    geom_point(show.legend=FALSE) + 
    ggtitle(assay) + scale_x_log10() + scale_y_log10()
}
plotProps(cla)

# normalization using edgeR
dds <- DGEList(assay(cla), group=cla$group)
dds <- calcNormFactors(dds)
assays(cla)$normalized <- cpm(dds)/10000 # from CPM to normalized %
# we compare the proportions
plotProps(cla) | plotProps(cla, "normalized")
```

The differential abundance analysis:

```{r}
dds <- estimateDisp(dds)
(out1 <- as.data.frame(topTags(exactTest(dds), Inf)))
```


```{r}
# we plot the proportions for the top population
d <- cbind(as.data.frame(colData(cla)), t(assays(cla)$normalized))
ggplot(d, aes(group, Microglia, fill=group)) + geom_violin() + stat_summary()
```


## Using explicit composition approaches

[ANCOMBC](https://bioconductor.org/packages/release/bioc/html/ANCOMBC.html) was
designed for microbiome data, and explictly models the effect of changes on some
subpopulations on the observed frequencies of others.

```{r}
# we first need to build the object as ANCOMBC expects it:
ps <- phyloseq(otu_table(assay(cla), taxa_are_rows=TRUE),
               sample_data(as.data.frame(colData(cla))))
# we run the method:
out2 <- ancombc(ps, formula="group")
# look at the results:
(out2 <- setNames(do.call(cbind, out2$res), names(out2$res)))
```

## Comparing the two

```{r, fig.width=8, fig.height=4}
out1$ANCOMBC.pval <- out2[row.names(out1),"p_val"]
out1$ANCOMBC.beta <- out2[row.names(out1),"beta"]
p1 <- ggplot(out1, aes(logFC, ANCOMBC.beta, size=-log10(ANCOMBC.pval))) + 
  geom_abline(slope=1, colour="grey") + geom_point() +
  labs(x="edgeR logFC", y="ANCOMBC beta", size="log10(pval)")
p2 <- ggplot(out1, aes((PValue), (ANCOMBC.pval))) + scale_x_log10() + scale_y_log10() +
  geom_abline(slope=1, colour="grey") + geom_point() +
  labs(x="edgeR PValue", y="ANCOMBC PValue")
p1 | p2
```


See [DA_exercise.Rmd](DA_exercise.Rmd) !

