---
title: "Continuum DA exercise"
author: "Mark"
date: "4/20/2022"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r}
suppressPackageStartupMessages({
  library(SingleCellExperiment)
  library(miloR) # differential expression analysis
  library(ggplot2) # for plotting
  library(patchwork) # for combining plots
  library(scater)
})
```

We read in the `SingleCellExperiment` object, do some clustering and interrogate some basic properties:

```{r}
# sce <- readRDS("/data/day1_differential_analysis/sce_continuum_exercise_public.rds")
sce <- readRDS("day1_differential_analysis/sce_continuum_exercise_public.rds")

plotUMAP(sce, colour_by="condition", shape_by="sample")

```


## Exercise

Let's try to find the subpopulations that are differentially abundant between condition A and B (3 replicates of each condition) using the methods discussed so far (one is very obvious from the low dimensional project). For example, one could even start with something straightforward: cluster cells into a reasonable resolution, extract a count table of those clusters across all samples and manually perform a DA analysis with your favourite tool. Following that, perhaps a miloR analysis could be tested or the inference of lineages (e.g., slingshot) and computing the (relative) densities along the lineage.
